Supplementary Material (SM)

### 2020-04-18 .A White Noise Approach To Evolutionary Ecology/SM.Rmd
split main text and SM

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued stochastic processes. This approach allows us to compute the full implications of demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a quantitative genetic model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests absolute values of linear selection gradients are negatively correlated with competition coefficients when the ratio of interspecific trait variation to intraspecific trait variation is approximately less than $1.25$ and positively correlated otherwise. The inverse relationship holds for quadratic selection gradients except with threshold diversity ratio equal to unity instead of $1.25$. Hence, this approach contributes to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of stochastic models that underlie rigorous investigations of the relationship between feedbacks of biological processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document: 
    extra_dependencies:
      - float
      - flafter
    fig_caption: yes
    fig_height: 3
    fig_width: 4
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{mathtools}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

@@

# Introduction

Our goal in this manuscript is to develop a rigorous, but accessible approach to synthesize the stochastic dynamics of abundance, mean trait and heritable variation in biological populations for the study of theoretical evolutionary ecology. A primary aim of theoretical evolutionary ecology is the development of mathematical approaches to describe the evolution of populations and their interactions with both the biotic and abiotic environments in which they are embedded. Given this consideration, a natural scope for such an approach centers on quantifying the abundance dynamics of populations and the evolution of traits mediating their interactions as functions of relevant abiotic factors. Although taking into account abundance, phenotype and environment provides the basis for a partial understanding of the complex nature of biological communities, a deeper understanding must account for the effects of dispersal and the phylogeographic history of interacting lineages [@Kraft2007; @Hickerson2010; @Manceau2016; @markmcpeek2017] along with the genetic basis of ecologically relevant traits [@jeffreyconner2004; @FUSSMANN2007] and feedbacks between populations and the biogeochemical cycles they ultimately depend on [@michelloreau2010; @9781107648258]. It is therefore ideal that the development of any such mathematical approach anticipates extensions to account for these important factors shaping ecological communities, especially as empirical and conceptual work in these directions continues to grow [@AbdalaRoberts2014; @Klzsch2015; @Crutsinger2015; @Fitzpatrick2015; @Fitzpatrick2017; @Marx2017; @Rudman2017; @Skovmand2018; @VanNuland2019; @Harmon2019]. Furthermore, the approach would benefit from a stochastic component to capture the chance nature of biological reality [@Lande2003; @DeMeester2018; @Mubayi2019] and serve as a basis for the construction of statistical methods that measure evolutionary and ecological processes occuring in the wild. Such methods will tether theory to reality and allow for rigorous tests of hypotheses on the structure and behavior of ecological communities. In this paper we introduce a framework that establishes a formal connection between the continuous-time dynamics of abundance and quantitative traits in stochastically evolving populations. We then demonstrate the utility of our framework through the derivation and analysis of a model of diffuse coevolution and discuss how it can be extended to account for the details mentioned above.

Current theoretical approaches to synthesize evolution and ecology have capitalized on the fact that biological fitness plays a key role in determining both sets of dynamics. While correlation of fitness and genotype is the basis of evolution by natural selection, the mean fitness across all individuals in a population determines the growth, stasis or decline of abundance. In section \ref{deterministic} we review the mathematical formalization of this connection, which has been established in the contexts of population genetics [@9781932846126; @joanroughgarden1979], evolutionary game theory [@Hofbauer1998; @9780674023383; @Lion2018], quantitative genetics [@Lande1982; @Doebeli1996; @Lion2018] and a unifying framework for these three distinct approaches to evolutionary theory [@Champagnat2006] which includes our approach as a special case. We note here there are two primary differences between the results presented in this paper and those of Champagnat, Ferrière and Méléard [-@Champagnat2006]. First, instead of developing a unifying approach to evolutionary theory, we focus on developing a stochastic synthesis of population dynamics and quantitative genetics including specific expressions for the dynamics of abundance, mean trait and additive genetic variance. Second, although Champagnat, Ferrière and Méléard [-@Champagnat2006] provide rigorous proofs of results relevant to our approach, they do not communicate the nessecary tools we make use of for deriving the dynamics of abundance, mean trait and additive genetic variance. We aim to communicate these tools to a wide audience by striking a balance between intuitive appeal and mathematical rigor.

Although reviewing previously established theoretical results reveals a beautiful synthesis of evolution and population ecology, it also reveals a gap in theoretical approaches to incorporate the intrinsically random nature of populations. Specifically, in theoretical quantitative genetics the derivation of a population's response to random genetic drift is derived in discrete time under the assumption of constant effective population size using arguments based on properties of random samples [@Lande1976]. Though this approach conveniently mimics the formalism provided by the Wright-Fisher model of population genetics, real population sizes fluctuate over time. Furthermore, since these fluctuations are themselves stochastic, it seems natural to derive expressions for the evolutionary response to demographic stochasticity and consider how the results relate to characterizations of random genetic drift. This can be done in continuous time for population genetic models without too much technical overhead, assuming a finite number of alleles [@Lande2009; @Parsons2010; @Gomulkiewicz2017]. However, for populations with a continuum of types, such as a quantitative trait, finding a formal approach to derive the evolutionary response to demographic stochasticity has remained a vexing mathematical challenge. In this paper we close this gap by combining the calculus of white noise with results on rescaled limits of branching Brownian motion processes (BBM) and stochastic partial differential equations (SPDE), which are stochastic analogs of deterministic partial differential equations (PDE). Our goal has two components: 1) Establish a novel synthetic approach to theoretical evolutionary ecology that provides a formal connection between demographic stochasticity and random genetic drift in the context of quantitative traits. 2) Communicate some useful properties of space-time white noise, BBM and SPDE to as wide of audience as possible. With this goal in mind we will not provide a rigorous treatment of any of these deep subjects. Instead, we introduce a set of heuristics that only require the basic concepts of Reimann integration, partial differentiation and some exposure to Brownian motion and stochastic ordinary differential equations (SDE). For a concise introduction to SDE and Brownian motion, we recommend the primer by Evans [-@9781470410544]. Rigorous treatments of SPDE and rescaled limits of BBM can be found in Walsh [-@Walsh] and Dawson [-@dawson1993measure] respectively.

Through accomplishing these goals this work yields five main results. Firstly, we develop a concise, but accurate approach for understanding SPDE and provide heuristics to empower mathematical evolutionary ecologists to derive novel models that relax key assumptions commonly made in combining models of evolutionary and abundance dynamics. Second, we formally derive general SDE driven by demographic stochasticity that express the joint dynamics of abundance, mean trait and additive genetic variance for populations exhibiting a wide range of potential phenotypic distributions. Although these dynamics have been considered before while accounting for demographic stochasticity [@Polechov2015], to our knowledge this is the first time they have been rigorously derived from first principles. Third, we simplify these SDE by approximating the abundance density of phenotypes with a Gaussian curve and integrate the results with a model of imperfect inheritance to find formula that generalize classical quantitative genetic expressions. This includes an expression for the response of additive genetic variance to selection in terms of fitness gradients analogous to the expression for the response in mean phenotype found by Lande [-@Lande1976]. Fourth, following classical niche theory we provide a rigorous mechanistic derivation of fitness for populations experiencing resource competition and abiotic stabilizing selection using diffusion limits of measure-valued branching processes. This calculation provides a demonstration for performing such diffusion limits that can be applied in other contexts. Fifth, we explore implications of the resulting model of diffuse coevolution by investigating the relationship between selection gradients and competition coefficients, discovering the conditions under which these quantities positively or negatively covary.

We now summarize the basic structure of the paper. To provide motivation for the stochastic equations developed later and background for our model of coevolution, we begin with \S\ref{deterministic} by briefly summarizing derivations of deterministic evolutionary and demographic dynamics. Starting with a PDE (which we refer to as the Deterministic Asexual Gaussian allelic model with Abundance dynamics, abbreviated DAGA), we arrive at a general set of ordinary differential equations modelling the dynamics of abundance, trait mean and trait variance. We discuss the relationship between DAGA and previous models of phenotypic evolution that have been employed in theoretical quantitative genetics. In \S\ref{wnc} we discuss how a diffusion limit of a spatially structured branching process (i.e., a measure-valued branching process) leads to a SPDE that provides a natural stochastic generalization of DAGA. We refer to this SPDE as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA). The diffusion limit in turn provides a rigorous method for constructing fitness functions used in models of evolutionary ecology. However, before we discuss measure-valued braching processes and their diffusion limits, we find it necessary to review some fundamental results on the calculus of white noise and introduce a set of mathematical tools based on these results. In SM \S\ref{SDE_DERIV} we employ these tools to derive a system of SDE generalizing our deterministic results to account for demographic stochasticity. We find the effect of demographic stochasticity on the evolution of mean trait and phenotypic variance characterizes the process of random genetic drift in continuous time. However, although biologically insightful, these equations remain difficult to analyze and implement numerically. In \S\ref{equations} we use an assumption of normally distributed trait values to simplify these expressions into tractable formula that express the evolutionary response to selection in terms of fitness gradients. We then account for the constraint of adaptive evolution on the amount of heritable variation available in a population by extending these results via a model of imperfect inheritance. The resulting equations coincide with classical results in quantitative genetics as a special case. In \S\ref{coev} we combine the derived equations of population dynamics with classical niche theory to formulate a model of coevolution across a guild of $S$ species participating in exploitative competition along a common resource continuum. In SM \S\ref{diffuse} we apply a result on rescaled limits of BBM to provide a rigorous derivation. In \S\ref{dynamics} we illustrate our model of coevolution through numerical integration for $S=100$ species, tracking the dynamics of traits and abundances under scenarios of weak and strong competition. We include an account of the natural history of the simulated system and discuss the significance of demographic stochasticity for structuring ecological communities. In \S\ref{ecoevo} we provide expressions for selection gradients and competition coefficients implied by our model and use these expressions to investigate the relationship between the degree of competition and strength of coevolution. Finally, \S\ref{conclusion} concludes with a summary of accomplishments, a few remarks on the limits of this approach and future directions to incorporate more explicitly the effects of small populations, the genetic architecture of quantitative traits, feedbacks with ecosystem processes and the macroevolutionary history of interacting lineages.

# The framework

At the core of our approach is a stochastic analog of the replicator equation with mutation in continuous time and phenotypic space [@Taylor1978; @Schuster1983]. From this stochastic replicator-mutator equation we derive a system of SDE for the dynamics of abundance, mean trait and additive genetic variance of a population. Hence, our approach develops a quantitative genetic theory of evolutionary ecology. A popular alternative to quantitative genetics is the theory of adaptive dynamics. As demonstrated by Page and Nowak [-@PAGE2002] and Champagnat, Ferrière and Méléard [-@Champagnat2006], the canonical equation of adaptive dynamics can be derived from the replicator-mutator equation. Thus, one could start from the atomic roots of our approach and pursue a stochastic adaptive dynamic theory instead. We choose the former in anticipation of an extension of our approach that explicitly models the genetic details of populations.

In this section we review the derivations of the replicator-mutator equation and trait dynamics from abundance dynamics and extend these formula along with related results to the stochastic case. The results established in this section provide the framework from which larger scale ecological stuctures, such as species abundance distributions and interaction networks, can be computed.

## Deterministic dynamics \label{deterministic}

Our review begins by considering the dynamics of an asexually reproducing population in a homogeneous environment. For simplicity, we first assume individuals are haploid and carry one of $K$ alleles each with a different fitness expressed as growth rate before introducing a model involving a quantitative trait. Under these assumptions, the derivation of the evolution of allele frequencies due to natural selection can be derived from expressions of exponential growth. This, and a few related approaches, have been provided by Crow and Kimura [-@9781932846126, \S 5.3]. Specifically, denoting $\nu_i$ the abundance of individuals with allele $i$ and $m_i$ the growth rate of allele $i$ [called the Malthusian parameter in @9781932846126], we have
\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation}
Starting from this model, we get the total abundance of the population as $N=\sum_{i=1}^K\nu_i$, the frequency of allele $i$ as $p_i=\nu_i/N$ and the mean fitness of the population as $\bar m=\sum_{i=1}^Kp_im_i$. Hence, we can employ some elementary calculus to derive the dynamics of abundance $dN/dt$ and the dynamics of allele frequencies $dp_1/dt,\dots,dp_K/dt$ as
\begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation}
\begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation}
Two important observations of these equations include (i) mean fitness $\bar m$ determines the abundance dynamics of the entire population and (ii) allele $i$ will increase (decrease) in frequency if $m_i>\bar m$ ($<\bar m$). Equation (\ref{popgen}) is known in the field of evolutionary game theory as the replicator equation [@Hofbauer1998; @9780674023383; @Lion2018; @Taylor1978; @Schuster1983]. Instead of being explicitly focused on alleles, the replicator equation describes the fluctuations of relative abundances of various _types_ in a population in terms of the vital rates of each type. Using a matrix of transition rates between differing types, it is straight-forward to extend the replicator equation to include mutation, which is known as the replicator-mutator equation [@9780674023383].

Inspired by equations (\ref{first})-(\ref{popgen}), we derive an analog of the replicator-mutator equation for a continuum of types (that is, for a quantitative trait). In particular, we model a continuously reproducing population with trait values $x\in\mathbb{R}$ and an abundance density $\nu(x,t)$ that represents the amount of individuals in the population with trait value $x$ at time $t$. Hence, the abundance density satisfies $N(t)=\int_{-\infty}^{+\infty}\nu(x,t)dx$ and $p(x,t)=\nu(x,t)/N(t)$ is the relative density of trait $x$ which we also refer to as the phenotypic distribution.

To stay within the realm of biological plausibility we require a set technical assumptions. First, we assume the initial abundance density is continuous, non-negative, integrable and has finite trait mean and variance. That is, we assume $\nu(x,0)$ is continuous in $x$, satisfies $\nu(x,0)\geq0$ for all $x\in\mathbb{R}$ and
\begin{equation}
N(0)=\int_{-\infty}^{+\infty}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{-\infty}^{+\infty}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{-\infty}^{+\infty}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(h,x)$ that is differentiable with respect to both arguments and is bounded above by some $r\in\mathbb{R}$ for all $x\in\mathbb{R}$ and $h\geq0$. We also assume $(1+h)C\leq m(h,x)$ for some constant $C>0$ so that $m(h,x)$ is linear or more than linear in $h$. We denote by $K$ an operator that takes functions such as $\nu(x,t)$ as its argument and returns other functions that may still depend on the spatial variable $x$ and the temporal variable $t$. For our application we will be focused on operators $K$ that can be written as $(K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x-y)\nu(y,t)dy$ for some non-negative and bounded function $\kappa$. We set $h=(K\nu)(x,t)$ to account for nonlocal effects of the abundance density $\nu(x,t)$ on the fitness of individuals with trait value $x$ [@Volpert2014]. As an example of a nonlocal effect, consider a population experiencing intraspecific competition for a common resource such that the competitive interactions are not mediated by any set of traits. In this case all individuals in the population effect the fitness of one another even though they may exhibit a diverse range of trait values. The term nonlocal comes from interpreting the trait value as spatial location. From here on we abbreviate $m(h,x)=m\big((K\nu)(x,t),x\big)$ to just $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. With these technicalities addressed we model the demographic dynamics of a population and the dynamics of a quantitative character simultaneously by the PDE
\begin{equation}\label{eq1}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)
\end{equation}
with the initial condition $\nu(x,0)$ described above. Equation (\ref{eq1}) is a semilinear parabolic PDE and is referred to as a scalar reaction-diffusion equation [@zheng2004nonlinear; @lawrenceevans2010]. To ensure existence and uniqueness of solutions to (\ref{eq1}), we further assume that for each $M>0$ there exists a constant $L_M>0$ depending on $M$ such that $\int_\mathbb{R}\nu_1(x)dx,\int_\mathbb{R}\nu_2(x)dx\leq M$ implies
\begin{equation}\label{local_lipschitz}
\int_\mathbb{R}\left|m(\nu_1,x)\nu_1(x)-m(\nu_2,x)\nu_2(x)\right|dx\leq L_M\int_\mathbb{R}|\nu_1(x)-\nu_2(x)|dx.
\end{equation}
In SM \S\ref{finite} we combine the above assumptions on growth rate, mutation and initial condition to prove that solutions to (\ref{eq1}) satisfy $N(t),|\bar x(t)|,\sigma^2(t)<+\infty$ for all $t\geq0$.

Interpreting the trait value $x$ as spatial location and ignoring nonlocal effects, which can be satisfied by setting $K\nu=\nu$, this model of spatially distributed population dynamics has been intensely studied [for a review, see @Cantrell2004]. When $\mu=0$, equation (\ref{eq1}) can be seen as an analog of equation (\ref{first}) for a continuum of types. By assuming mutation acts as diffusion, the effect of mutation causes $\nu(x,t)$ to flatten out over time. In fact, if the growth rate is constant across $x$, then this model of mutation will cause $\nu(x,t)$ to converge to a flat line as $t\to\infty$.

Although clearly an idealized representation of biological reality, this model is sufficiently general to capture a large class of dynamics including density dependent growth and frequency dependent selection. As an example, logistic growth combined with quadratic stabilizing selection can be captured using the growth rate 
\begin{equation}\label{special_m}
m(\nu,x)=r-\frac{a}{2}(\theta-x)^2-c\int_{-\infty}^{+\infty}\nu(y,t)dy
\end{equation}
where $r\in\mathbb{R}$ is the intrinsic rate of growth in the absence of abiotic stabilizing selection, $a>0$ is the strength of abiotic stabilizing selection around the phenotypic optimum $\theta\in\mathbb{R}$, $c>0$ captures the sensitivity of fitness to intraspecific competition and $\kappa(x-y)=1$ so that competitive interactions cause the same reduction in fitness regardless of trait value. This exemplary fitness function has a few convenient properties. First, the effect of competition induces a local carrying capacity on the population, leading to a finite equilibrium abundance over bounded subsets of $\mathbb{R}$. Second, abiotic selection prevents the abundance density from diffusing too far from the abiotic optimum, leading to a finite equilibrial abundance across all of $\mathbb{R}$, $\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})$, finite equilibrial mean trait $\lim_{t\to\infty}\bar x(t)=\theta$ and finite equilibrial phenotypic variance $\lim_{t\to\infty}\sigma^2(t)=\sqrt{\frac{\mu}{a}}$, so long as $\bar x(0)\in\mathbb{R}$, $\sigma^2(0)\in[0,+\infty)$, $N(0)\in(0,+\infty)$ and $r>\frac{1}{2}\sqrt{a\mu}>0$. We prove these statements in SM \S\ref{equilib}. Note in particular the coincidence of equilibrial phenotypic variance predicted by this model and that predicted by the classic Gaussian allelic model of quantitative genetics [@Lande1975; @9780471986539; @Johnson2005]. In \S\ref{coev} and SM \S\ref{diffuse} we find this fitness function is closely related to a fitness function derived from niche theory that we use to model coevolution driven by resource competition.

To derive a replicator-mutator equation from equation (\ref{eq1}), we employ the chain rule from calculus. Writing $\bar m(t)=\int_{-\infty}^{+\infty}m(\nu,x)p(x,t)dx$ for the mean fitness, we have

\begin{multline}\label{N_det}
\frac{d}{dt}N(t)=\frac{d}{dt}\int_{-\infty}^{+\infty}\nu(x,t)dx=\int_{-\infty}^{+\infty}\frac{\partial}{\partial t}\nu(x,t)dx \\
=\int_{-\infty}^{+\infty} m(\nu,x)\nu(x,t)dx+\int_{-\infty}^{+\infty}\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)dx \\
=N(t)\int_{-\infty}^{+\infty} m(\nu,x)p(x,t)dx=\bar m(t) N(t).
\end{multline}
Using our assumptions on mutation and rate of growth, we show in SM \S\ref{finite} $\nu(x,t)$ is twice differentiable with respect to $x$ and $\int_{-\infty}^{+\infty}\nu(x,t)dx<\infty$ for all $t\geq0$. This implies that we are justified in swapping the order of differentiation and integration and the result $\int_{-\infty}^{+\infty}\frac{\partial^2}{\partial x^2}\nu(x,t)dx=0$ can be derived from the fundamental theorem of calculus. Biological reasoning agrees with this latter result since mutation neither creates nor destroys individuals, but merely changes their type from their parental type. Taking the same approach, we derive the dynamics of the phenotypic distribution $p(x,t)$ in response to selection and muation as

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=\Big(m(\nu,x)-\bar m(t)\Big) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}
This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which, for sexually reproducing populations, has been referred to as the Gaussian allelic approximation in reference to the distribution of mutational effects at loci in a genome on the values of traits [@Lande1975; but see also @9780471986539; along with @Johnson2005], the infinitesimal model in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect [@Barton2017] and the Gaussian descendants approximation in reference to offspring trait values being normally distributed around their parental values [@Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of Kimura's continuum-of-alleles model. We return to these concepts in \S\ref{inheritance} where we review a simple model of inheritance that leads us to an expression for the stochastic evolution of additive genetic variance. To distinguish this model from previous models of phenotypic evolution we refer to PDE (\ref{eq1}) from which (\ref{p_PDE}) was derived as the Deterministic Asexual Gaussian allelic model with Abundance dynamics (abbreviated DAGA). Later, we will extend this model to include the effects of demographic stochasticity, which we refer to as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA).

The covariance of fitness and phenotype across the population is defined as
\begin{equation}
\mathrm{Cov}_t\Big(m(\nu,x),x\Big)=\int_{-\infty}^{+\infty}\Big(m(\nu,x)-\bar m(t)\Big)\Big(x-\bar x(t)\Big)p(x,t)dx.
\end{equation}
Hence, the dynamics of the mean trait $\bar x(t)$ can be derived as

\begin{multline}\label{detmean}
\frac{d}{dt}\bar x(t)=\frac{d}{dt}\int_{-\infty}^{+\infty} x p(x,t)dx=\int_{-\infty}^{+\infty} x\frac{\partial}{\partial t} p(x,t)dx \\
=\int_{-\infty}^{+\infty} x\big(m(\nu,x)-\bar m(t)\big) p(x,t)+x\frac{\mu}{2}\frac{\partial^2}{\partial x^2} p(x,t)dx
=\mathrm{Cov}_t\Big(m(\nu,x),x\Big).
\end{multline}

Equation (\ref{detmean}) is a continuous time analog of the well known Robertson-Price equation without transmission bias [@Robertson1966; @PRICE1970; @FRANK2012; @Queller2017; @Lion2018]. Whether or not the covariance of fitness and phenotype creates change in $\bar x$ to maximize mean fitness $\bar m$ depends on the degree to which selection is frequency dependent [@Lande1976]. Since this change is driven by a covariance with respect to phenotypic diversity, the response in mean trait to selection is mediated by the phenotypic variance. In particular, when $\sigma^2=0$, $\bar x$ will not respond to selection. The result $\int_{-\infty}^{+\infty} x\frac{\partial^2}{\partial x^2} p(x,t)dx=0$ can be found by applying integration by parts. Following the approach taken to calculate the evolution of $\bar x$, we find the response of phenotypic variation to this model of selection and mutation is
\begin{equation}\label{vardyn_det}
\frac{d}{dt}\sigma^2(t)=\mathrm{Cov}_t\Big(m(\nu,x),(x-\bar x)^2\Big)+\mu.
\end{equation}
For the sake of space we relegate the derivation of $d\sigma^2/dt$ to SM \S\ref{var_deriv}. In the absence of mutation equation (\ref{vardyn_det}) agrees with the result derived by Lion [-@Lion2018] for discrete phenotypes. From a statistical perspective, if we think of $(x-\bar x)^2$ as a square error, then in analogy to the dynamics of the mean trait, we see that the response in $\sigma^2$ to selection can be expressed as a covariance of fitness and square error, which is defined in analogy to $\mathrm{Cov}_t(m(\nu,x),x)$. Just as for the evolution of $\bar x(t)$, this covariance also creates change in $\sigma^2$ that can either increase or decrease mean fitness $\bar m$, depending on whether or not individual fitness depends on the distribution of traits across the entire population. The effect of selection on phenotypic variance can be positive or negative depending on whether selection is stabilizing or disruptive.

In SM \S\ref{SDE_DERIV}, we extend these results to include the effects of demographic stochasticity. The key realization in this direction is that a rigorous derivation of a SPDE that naturally generalizes DAGA and accounts for ecological interactions has been provided by Li [-@zeng1998absolute] and slightly generalized by Champagnat, Ferriére and Méléard [-@Champagnat2006]. To understand how this SPDE was derived, we provide an informal discussion of measure-valued branching processes (which treat populations as sets of discrete individuals) and their diffusion limits, the so-called superprocesses (which treat population size as a continuously varying number). Similar to the approach taken by Champagnat, Ferriére and Méléard [-@Champagnat2006], we begin with a BBM that models populations as discrete sets of reproducing individuals whose vital rates depend on their trait value as well as the state of the entire population. Taking a large population size limit and keeping our assumption of single dimensional traits, we employ a pair of classical results that show, under the appropriate rescaling in time, phenotypic space and population density, a sequence of rescaled BBM converges to a limiting process that can be characterized by a SPDE [@meleard1992interacting; @Mlard1993; @zeng1998absolute; @Champagnat2006]. The limiting processes of rescaled BBM have been referred to as measure-valued Markov processes [@dawson1993measure] or superprocesses [@alisonetheridge2000]. Under the simplifying assumptions inherited from our treatment of deterministic dynamics and the additional assumption that the magnitude of demographic stochasticity is independent of trait values, we obtain as a special case a relatively simple expression for an SPDE that generalizes DAGA. The simplicity of our special case allows us to use properties of space-time white noise processes to derive a set of SDE that generalize equations (\ref{N_det}), (\ref{detmean}) and (\ref{vardyn_det}) to include the effects of demographic stochasticity. Classical expressions for the effects of random genetic drift on the evolution of mean traits are obtained as a further special case.

In the following section we provide the necessary mathematical tools needed to derive SDE from SPDE. Since our aim is to present this material to as wide of audience as possible, our approach deviates from standard definitions to remove the need for a detailed technical treatment. In addition to the notions of Reimann integration and partial differentiation already employed, the reader will only need some elementary probability and an intuitive understanding of SDE, including Brownian motion. Because space-time white noise, denoted by $\dot W(x,t)$, appears in the SPDE characterizing diffusion limits of BBM, we begin by defining $\dot W(x,t)$ and illustrating its relevant properties including a set of heuristics for performing calculations. Treating only the simplist of cases, we then provide a brief review of BBM, their diffusion limits and the SPDE that characterize them. For those not interested in the white noise calculus or superprocesses and would rather jump straight into more biologically relevant results, we recommend skipping to \S\ref{equations}.

## White noise calculus and superprocesses \label{wnc}

### Introduction to white noise

One can think of white noise as the static seen on old television sets or infinitely detailed random dust spread across both time and space. From a more mathematical, yet still informal perspective, white noise can be thought of as a stochastic process. That is, we can picture white noise as a collection of random variables indexed by time and possibly space. In relation to Brownian motion, denoted by $W$, white noise can be interpreted of as the derivative of Brownian motion with respect to time, denoted $\dot W$. Since Brownian motion can be thought to take infinitesimally small Gaussian distributed jumps at each time point, this leads to the conceptualization of white noise as a collection of independent Gaussian distributed random variables. Figure \ref{wn} illustrates realizations of this conceptualized white noise in one (left) and two (right) dimensions.

```{r, include=F, eval=F}
require("ggplot2")
require("ggthemes")
require("plot3D")
require("gridExtra")

num1 = 1e4
num2 = 1e4

xi_1 = rnorm(num1)
xi_2 = rnorm(num2)

x1 = seq(0,1,length.out = num1)
x2 = seq(0,1,length.out = sqrt(num2))

M = expand.grid(x=x2,y=x2)
M$value = xi_2

df_1 = data.frame(xi=xi_1,x=x1)

wn1d = ggplot(df_1)+geom_point(aes(x=x,y=xi),size=0.05)+theme_minimal()+xlab("Time")+ylab("Value")
wn2d = ggplot(M, aes(x, y, color = value))+scale_color_gradient2(low="black",mid="grey50", high="white")+geom_point(size=0.001, shape=22, show.legend = FALSE)+theme_minimal()+xlab("Time")+ylab("Space")

wn = grid.arrange(wn1d,wn2d,nrow=1)

ggsave("~/Gits/branching.brownian.motion.and.spde/.A White Noise Approach To Evolutionary Ecology/wn.png",wn,width=5,height=2.5)
```

```{r, echo=F, fig.align='center', fig.cap='\\label{wn}Approximations of sample paths of temporal white noise (left) and space-time white noise (right) with brightness scaled to value.'}
knitr::include_graphics('wn.png')
```

However, it turns out that realizations of white noise do not exist as functions in the classical sense. Indeed, since Brownian motion is nowhere differentiable with respect to time, white noise cannot be formally understood as a time derivative. Thus our notation $\dot W$ is only meant to aid intuition and not be taken as formal. A formal understanding is possible by considering white noise as a _measure_-valued process [@Dawson1975; @Walsh] or as a _generalized_ process that acts on classically defined functions or stochastic processes to return either random variables or stochastic processes [@Krylov1981; @DaPrato2014]. Since a measure-valued process can be defined from a generalized process and a generalized process can be defined from a measure-valued process, the distinction between the two is more or less a matter of perspective. However, we find the perspective of white-noise as a generalized process to be a more efficient route for developing heuristics to help with some routine calculations involved with deriving SDE from SPDE. Hence, the notion of a generalized process provides the general idea implemented here. Although the treatments of SPDE provided by Krylov and Rozovskii [-@Krylov1981] and Da Prato and Zabczyk [-@DaPrato2014] extend the theory of SDE to formally treat SPDE in a general and elegant fashion, they require the navigation of an enourmous amount of technical definitions and detailed proofs. To extract some particularly useful results from this theory relevant to our goal of synthesizing the stochastic dynamics of biological populations, we provide a streamlined approach by capitolizing on the solid ground these authors have established. For instance, instead of rigorously proving properties of white noise, we simply define them to be so, taking solice in the fact that the technical details have been worked out elsewhere. In SM \S\ref{DPZ14} we show how our informal treatment is related to the rigorous treatment provided by Da Prato and Zabczyk [-@DaPrato2014].

Before diving in, we shed a bit of light on the idea of a generalized process. A generalized process is the stochastic analog of a generalized function, such as the Dirac delta function $\delta$. Often one sees $\delta$ defined as a function satisfying the properties $\delta(x)=0$ for $x\neq0$ and $\int_{-\infty}^{+\infty}\delta(x)dx=1$. However, since there is no function that satisfies these properties, we refer to $\delta$ as a generalized function. An alternative definition of $\delta$ considers its action on classically defined functions $f$. In particular, $\delta(f)=f(0)$, which can be heuristically written as $\int_{-\infty}^{+\infty}f(x)\delta(x)dx=f(0)$. Similarly, other generalized functions can be defined by their action on classically defined functions. Then, just as a generalized function operates on classical functions to return a value, a generalized process acts on a set of functions (or processes) to return a random variable (or a classically defined stochastic process). For a brief primer on the theory of generalized functions, see the addendum to chapter 3 of Kolmogorov and Fomin [-@kolmogorov1999elements].

### Definition and basic properties of white noise \label{wnc_intro}

>>>>

---
title: "Supplementary Material (SM)"
bibliography: ref.bib
output:
  pdf_document: 
    extra_dependencies:
      - float
      - flafter
    fig_caption: yes
    fig_height: 3
    fig_width: 4
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{xr}
  - \usepackage{mathtools}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
  - \externaldocument{draft}  
editor_options: 
  chunk_output_type: console
---

@@
